Planning to Save Energy: How Information Format Affects Accuracy

### 2025-01-04 docs/misc/ordinal.qmd
writing updates from last week I forgot to push

pacman::p_load(dplyr,purrr,tidyr,here,tibble,stringr,brms,rstan,bayestestR,emmeans,tidybayes,modelsummary,
               ggplot2,gt,knitr,kableExtra,ggh4x,lme4,flextable,pander)
options(digits=2, scipen=999, dplyr.summarise.inform=FALSE)
walk(c("fun_plot"), ~ source(here::here(paste0("scripts/", .x, ".R"))))
theme_set(theme_nice())
options(brms.backend="cmdstanr",mc.cores=4)

s1 <- readRDS(here::here("data/s1_processed.rds")) |> filter(!(id %in% readRDS(here::here("data/s1_discrep_ids.rds"))))
s2_long <- readRDS(here::here("data/s2_processed.rds")) |> filter(!(id %in% readRDS(here::here("data/s2_discrep_ids.rds"))))
```

@@

ordinal_model_s2_logit <- brm(
  accuracy_level ~ refClass +rounded+pct_goal + (1|id)+ (1|state),
  data = s2_agg1,
  family = cumulative("logit"),
  cores = 4,
  iter = 2000,
  control = list(adapt_delta = 0.99), # Recommended for ordinal models
  prior = c(prior(normal(0, 2), class = "Intercept"),  # Priors for thresholds
            prior(normal(0, 2), class = "b")), # Priors for predictors
  file = paste0(here::here("data/model_cache",'s2_logit_add.rds')) # Cache for efficiency
)
summary(ordinal_model_s2_logit)
# Regression Coefficients:
#                    Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# Intercept[1]          -2.13      0.64    -3.39    -0.86 1.00      639     1328
# Intercept[2]          -0.62      0.63    -1.89     0.63 1.00      658     1366
# Intercept[3]           3.15      0.65     1.88     4.42 1.00      717     1638
# refClassPercentage     0.83      0.75    -0.64     2.33 1.00      452     1253
# refClassUSD            1.87      0.79     0.31     3.35 1.00      508     1260
# roundedRounded        -0.66      0.18    -1.01    -0.31 1.00     4276     3095
# pct_goal15%           -0.44      0.18    -0.79    -0.10 1.00     6263     3109

pred_summary_s2 <- ordinal_model_s2_logit %>%
  epred_draws(newdata = s2_agg1, re_formula = NA) %>%
  group_by(refClass, rounded, pct_goal, .category) %>%
  summarise(
    mean_prob = mean(.epred),
    lower_ci = quantile(.epred, 0.025),
    upper_ci = quantile(.epred, 0.975),
    .groups = "drop"
  )

>>>>

pacman::p_load(dplyr,purrr,tidyr,here,tibble,stringr,brms,rstan,bayestestR,emmeans,tidybayes,modelsummary,
               ggplot2,gt,knitr,kableExtra,ggh4x,lme4,flextable,pander,marginaleffects,ggstance)
options(digits=2, scipen=999, dplyr.summarise.inform=FALSE)
walk(c("fun_plot"), ~ source(here::here(paste0("scripts/", .x, ".R"))))
theme_set(theme_nice())
options(brms.backend="cmdstanr",mc.cores=4)

# s1 <- readRDS(here::here("data/s1_processed.rds")) |> filter(!(id %in% readRDS(here::here("data/s1_discrep_ids.rds"))))
# s2_long <- readRDS(here::here("data/s2_processed.rds")) |> filter(!(id %in% readRDS(here::here("data/s2_discrep_ids.rds"))))

s1 <- readRDS(here::here("data/s1_processed.rds")) |> 
  filter(!(id %in% readRDS(here::here("data/s1_discrep_ids.rds")))) |> 
  filter(!(id %in% readRDS(here::here("data/s1_grp_outlier_ids.rds")))) |>
  mutate(refClass = factor(refClass, levels=c("kWh","Percentage","USD")))

s2_long <- readRDS(here::here("data/s2_processed.rds")) |> 
  filter(!(id %in% readRDS(here::here("data/s2_discrep_ids.rds")))) |> 
  filter(!(id %in% readRDS(here::here("data/s2_grp_outlier_ids.rds"))) ) |>
  mutate(refClass = factor(refClass, levels=c("kWh","Percentage","USD")))

# s2 |> group_by(appliance,family,state,state_avg) |> select(appliance,family_prior_yr=family,state,state_avg) |> 
#   slice_head() |> print(n=30) |> pander::pandoc.table(split.table=Inf)
# 
# s2 |> group_by(appliance,family,state,state_avg) |> select(appliance,family_prior_yr=family,state,state_avg) |>
#   arrange(state,appliance) |>
#   slice_head() |> print(n=30) |> kable()
# 
# s2 |> head() |> ungroup() |>select(-edu,-els,-block) |> kable()
```

@@

ordinal_model_s2_logit <- brm(
  accuracy_level ~ refClass +rounded+pct_goal + (1|id)+ (1|state),
  data = s2_agg,
  family = cumulative("logit"),
  cores = 4,
  iter = 2000,
  control = list(adapt_delta = 0.99), # Recommended for ordinal models
  prior = c(prior(normal(0, 2), class = "Intercept"),  # Priors for thresholds
            prior(normal(0, 2), class = "b")), # Priors for predictors
  file = paste0(here::here("data/model_cache",'s2_logit_add.rds')) # Cache for efficiency
)
summary(ordinal_model_s2_logit)
# Regression Coefficients:
#                    Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
# Intercept[1]          -2.13      0.64    -3.39    -0.86 1.00      639     1328
# Intercept[2]          -0.62      0.63    -1.89     0.63 1.00      658     1366
# Intercept[3]           3.15      0.65     1.88     4.42 1.00      717     1638
# refClassPercentage     0.83      0.75    -0.64     2.33 1.00      452     1253
# refClassUSD            1.87      0.79     0.31     3.35 1.00      508     1260
# roundedRounded        -0.66      0.18    -1.01    -0.31 1.00     4276     3095
# pct_goal15%           -0.44      0.18    -0.79    -0.10 1.00     6263     3109

pred_summary_s2 <- ordinal_model_s2_logit %>%
  epred_draws(newdata = s2_agg, re_formula = NA) %>%
  group_by(refClass, rounded, pct_goal, .category) %>%
  summarise(
    mean_prob = mean(.epred),
    lower_ci = quantile(.epred, 0.025),
    upper_ci = quantile(.epred, 0.975),
    .groups = "drop"
  )

### 2025-01-04 manuscript/appendix.qmd
writing updates from last week I forgot to push

>>>>

# A tibble: 4,560 × 17
# Groups:   id, block [380]
   id     block refClass appliance pct_goal rounded state    calc       edu                         els   pct family state_avg plan  value change state_dif
   <chr>  <dbl> <fct>    <chr>     <chr>    <chr>   <chr>    <chr>      <fct>                     <dbl> <dbl>  <dbl>     <dbl> <fct> <dbl>  <dbl>     <dbl>
 1 s2_100     1 kWh      AC        10%      Rounded Colorado Calculator Highschool diploma or GED 0.875   0.1   1000       498 plan1   500   -500         2
 2 s2_100     1 kWh      AC        10%      Rounded Colorado Calculator Highschool diploma or GED 0.875   0.1   1000       498 plan2   300   -700      -198
 3 s2_100     1 kWh      HEAT      10%      Rounded Colorado Calculator Highschool diploma or GED 0.875   0.1  18000     16411 plan1 15000  -3000     -1411
 4 s2_100     1 kWh      HEAT      10%      Rounded Colorado Calculator Highschool diploma or GED 0.875   0.1  18000     16411 plan2 12000  -6000     -4411
 5 s2_100     1 kWh      WATER     10%      Rounded Colorado Calculator Highschool diploma or GED 0.875   0.1  12000      5832 plan1  6000  -6000       168
 6 s2_100     1 kWh      WATER     10%      Rounded Colorado Calculator Highschool diploma or GED 0.875   0.1  12000      5832 plan2  2000 -10000     -3832
 7 s2_100     1 kWh      REFRIG    10%      Rounded Colorado Calculator Highschool diploma or GED 0.875   0.1   1000      1142 plan1  2000   1000       858
 8 s2_100     1 kWh      REFRIG    10%      Rounded Colorado Calculator Highschool diploma or GED 0.875   0.1   1000      1142 plan2   500   -500      -642

# widen by appliance, with separate row for value, family, and stage_avg

s2 <- s2_long #|> select(-family,-state_avg,-edu,-change,-state_dif) |> pivot_wider(names_from = appliance, values_from = c(value)) 
s2s <- s2_long |> select(-family,-state_avg,-edu,-change,-value) |> pivot_wider(names_from = appliance, values_from = c(state_dif)) |> select(id,block,refClass,pct,plan,AC:OTHER)
# s2s
#    id     block refClass     pct plan     AC   HEAT WATER REFRIG OTHER
#    <chr>  <dbl> <fct>      <dbl> <fct> <dbl>  <dbl> <dbl>  <dbl> <dbl>
#  1 s2_100     1 kWh         0.1  plan1     2  -1411   168    858   348
#  2 s2_100     1 kWh         0.1  plan2  -198  -4411 -3832   -642 -6152
#  3 s2_100     2 kWh         0.15 plan1  1324     19  -139    320  -424
#  4 s2_100     2 kWh         0.15 plan2   324   -981 -1139   -318 -1424
#  5 s2_101     1 kWh         0.1  plan1   102   -411  2168   -142   398
#  6 s2_101     1 kWh         0.1  plan2   502  -1411   168   -142  1348
#  7 s2_101     2 kWh         0.15 plan1   751    901  -396    682 -1883
#  8 s2_101     2 kWh         0.15 plan2  2251    401   104    682 -1383
#  9 s2_102     1 Percentage  0.1  plan1   958 -12954 -5354    200 -5417
# 10 s2_102     1 Percentage  0.1  plan2   890 -15435 -4956   8544  3225

# clustering planning strategies - kmeans

    ‘
s2s |>  ungroup() |> select(-id,-block,-refClass,-pct,-plan) |> scale(center = TRUE, scale = TRUE) |> kmeans(centers = 3) 
s2s_clust <- s2s |>  ungroup() |> select(-id,-block,-refClass,-pct,-plan) |> scale(center = TRUE, scale = TRUE) |> kmeans(centers = 3) %>% .$cluster |> as.factor() 
s2s <- cbind(s2s,c=s2s_clust)

s2s |>    group_by(c) |>
  summarise(ac=mean(AC),heat=mean(HEAT),water=mean(WATER),refrig=mean(REFRIG),other=mean(OTHER)) |>
  pivot_longer(cols = c(ac:other), names_to = "Appliance", values_to = "Mean") |>
  ggplot(aes(x=Appliance,y=Mean,col=c)) + 
  geom_point() + 
  facet_wrap(~c) +
  theme_minimal()

s2s <- s2_long |> select(-family,-state_avg,-edu,-change,-state_dif) |> pivot_wider(names_from = appliance, values_from = c(value))  |> select(id,block,refClass,pct,plan,AC:OTHER)

s2s |>  ungroup() |> select(-id,-block,-refClass,-pct,-plan) |> scale(center = TRUE, scale = TRUE) |> kmeans(centers = 3)

s2s_clust <- s2s |>  ungroup() |> select(-id,-block,-refClass,-pct,-plan) |> scale(center = TRUE, scale = TRUE) |> kmeans(centers = 3) %>% .$cluster |> as.factor()

s2s <- cbind(s2s,c=s2s_clust) |> left_join(s2, join_by(id,refClass,block,plan,pct))
s2s |>  group_by(c) |>
  summarise(ac=mean(AC),heat=mean(HEAT),water=mean(WATER),refrig=mean(REFRIG),other=mean(OTHER)) |>
  pivot_longer(cols = c(ac:other), names_to = "Appliance", values_to = "Mean") |>
  ggplot(aes(x=Appliance,y=Mean,col=c)) + 
  geom_point() + 
  geom_point(aes(y=state_avg),color="black") +
  facet_wrap(~c) +
  theme_minimal()

s2 |> group_by(appliance,family,state,state_avg) |> select(appliance,family_prior_yr=family,state,state_avg)
```

### 2025-01-04 manuscript/manuscript.qmd
writing updates from last week I forgot to push

---
title: "Planning to Save Energy: How Information Format Affects Accuracy"
aliases:
  - /manuscript.html
format:
    wordcount-html: 
        #output-file: index.html
        wordcount-banner: true
        theme: 
            - zephyr
            - ../assets/custom.scss
        toc: true
        shift-heading-level-by: 1
        title-block-banner: "#000D51"
        include-in-header:
            - text: |
                <style type="text/css">
                h2.hide {
                    display: none;
                }
                </style>
    html: 
      echo: true
      code-fold: show
      output-file: manuscript-code.html
    hugo-md:
        include: false
        echo: false
        html-math-method: mathjax
        output-file: manuscript_hugo.md
        suppress-bibliography: true
    gfm:
        echo: true
        output-file: manuscript_gfm.md
        suppress-bibliography: true
    hikmah-manuscript-pdf:
        output-file: rep_study.pdf
        mainfont: "Linux Libertine O"
        mainfontoptions:
        - "Numbers=Proportional"
        - "Numbers=OldStyle"
        mathfont: "Libertinus Math"
        knitr:
            opts_chunk:
                dev: "cairo_pdf"
    #hikmah-manuscript-docx: default
    docx: default
format-links:   
    - text: HTML
      href: manuscript.html
      icon: file-minus
    - text: HTML (with code)
      href: manuscript-code.html
      icon: file-plus
    - format: hikmah-manuscript-pdf
      text: PDF version
      icon: file-pdf
    - docx
    - gfm
date: 2024-12-07
abstract: |
    Effective communication of energy consumption information is crucial for promoting residential energy conservation. This study investigates how different numerical representations of energy reduction goals influence consumers' ability to create accurate conservation plans. Across two experiments, we examined the impact of presenting energy information in kilowatt-hours (kWh), percentages, or U.S. dollars (USD) on planning accuracy. Participants completed a simulated household planning task in which they allocated energy usage across multiple appliances, with the goal presented in either kilowatt-hours (kWh), percentages, or monetary costs. Results across both experiments showed that presenting reduction goals in absolute units (kWh) led to significantly greater accuracy compared to percentage-based or monetary formats. Furthermore, we found that higher energy literacy was associated with more accurate planning. These findings demonstrate that absolute units (kWh) are more effective for communicating energy-saving goals, and highlight the potential value of educational interventions to improve consumer energy literacy.
---

@@

- highlight the significant contribution of residential energy consumption to carbon emissions and the potential for substantial reductions. Make point about urgency of climate change? (IPCC (2014) and EIA (2012) ).

- Electricity bills are a primary source of energy-use information for consumers and offer a promising avenue for enhancing communication about energy consumption [@fischerFeedbackHouseholdElectricity2008]

@@

Energy poverty continues to be a pervasive issue in the United States [@memmottSociodemographicDisparitiesEnergy2021]. This challenge partly arises from difficulties in converting information across numerical formats, impeding the development of precise energy reduction plans @reimerNumericCommunicationRisk2015. Prior research by @canfieldPerceptionsElectricityuseCommunications2017 demonstrated that presenting energy information in tabular formats enhances comprehension relative to graphs. Nevertheless, persistent difficulties remain in conveying energy data in formats that foster accurate consumer judgments and planning, as consumers often lack intuitive benchmarks for interpreting abstract units such as kilowatt-hours [@attariPublicPerceptionsEnergy2010].

@@

Furthermore, research suggests that natural frequencies and absolute numbers are generally easier for individuals to understand compared to percentages or probabilities [@gigerenzerHowImproveBayesian1995 @hoffrageCommunicatingStatisticalInformation2000]. In the context of energy conservation, using absolute units may facilitate more accurate planning and decision-making by aligning with intuitive cognitive processing.

@@

For each family scenario, the participants were shown a table containing the families utility usage from the prior year, alongside the state averages for each appliance category (both prior year usage and stage averages are always shown in kWh). For each scenario, participants were asked to create two possible action plans to achieve the target reduction in total household energy usage (see @fig-task). Depending on their reference class condition, the target reduction amount presented either in kilowatt-hours (kWh), as percentages of total household usage, or in U.S. dollars. In all conditions, the target reduction was equivalent to a 15% reduction in total household kWh.

@@

The experimental procedures in Experiment 2 are quite similar to those in Experiment 1. Experiment 2 employed a 2 (task goal: 10% vs. 15% reduction) x 2 (last year's usage: exact vs. rounded) within-subjects design, with a between-subjects manipulation of the reference class (USD vs. Percentage vs. kWh).. We recruited 206 participants from Amazon Mechanical Turk, but data from from 10 participants were corrupted due to experimenter error, leaving a final sample of 196 participants.

@@

We once again examained the effect of energy literacy on planning accuracy. A Bayesian linear regression model was fit with log-transformed absolute error as the outcome variable and energy literacy score as the predictor, controlling for random effects of participant and state: log_abs_error ~ els + (1|id) + (1|state). This revealed a significant negative relationship between energy literacy and log absolute error (Estimate = -3.21, 95% CI: -3.89 to -2.52), indicating that participants with higher energy literacy scores tended to have smaller deviations from the target reduction goal, and thus more accurate plans overall (@fig-s2-els).

@@

# General Discusion

>>>>

---
title: "Planning to Save Energy: How Information Format Affects Accuracy"
aliases:
  - /manuscript.html
format:
    wordcount-html: 
        #output-file: index.html
        wordcount-banner: true
        echo: false
        theme: 
            - zephyr
            - ../assets/custom.scss
        toc: true
        shift-heading-level-by: 1
        title-block-banner: "#000D51"
        include-in-header:
            - text: |
                <style type="text/css">
                h2.hide {
                    display: none;
                }
                </style>
    html: 
      echo: true
      code-fold: show
      output-file: manuscript-code.html
    hugo-md:
        include: false
        echo: false
        html-math-method: mathjax
        output-file: manuscript_hugo.md
        suppress-bibliography: false
    gfm:
        echo: true
        output-file: manuscript_gfm.md
        suppress-bibliography: true
    hikmah-manuscript-pdf:
        output-file: rep_study.pdf
        mainfont: "Linux Libertine O"
        mainfontoptions:
        - "Numbers=Proportional"
        - "Numbers=OldStyle"
        mathfont: "Libertinus Math"
        knitr:
            opts_chunk:
                dev: "cairo_pdf"
    #hikmah-manuscript-docx: default
    docx: default
format-links:   
    - text: HTML
      href: manuscript.html
      icon: file-minus
    - text: HTML (with code)
      href: manuscript-code.html
      icon: file-plus
    - format: hikmah-manuscript-pdf
      text: PDF version
      icon: file-pdf
    - docx
    - gfm
date: 2024-12-07
abstract: |
    Effective communication of energy consumption information is crucial for promoting residential energy conservation. This study investigates how different numerical representations of energy reduction goals influence consumers' ability to create accurate conservation plans. Across two experiments, we examined the impact of presenting energy information in kilowatt-hours (kWh), percentages, or U.S. dollars (USD) on planning accuracy. Participants completed a simulated household planning task in which they allocated energy usage across multiple appliances, with the goal presented in either kilowatt-hours (kWh), percentages, or monetary costs. Results across both experiments showed that presenting reduction goals in absolute units (kWh) led to significantly greater accuracy compared to percentage-based or monetary formats. Furthermore, we found that higher energy literacy was associated with more accurate planning. These findings demonstrate that absolute units (kWh) are more effective for communicating energy-saving goals, and highlight the potential value of educational interventions to improve consumer energy literacy.
---

@@

@@

Electricity bills are a frequent touchpoint for consumers seeking to understand their energy usage, yet effectively turning that billing data into practical steps can be demanding. Among low-income households, these challenges contribute to disproportionate rates of energy insecurity [@memmottSociodemographicDisparitiesEnergy2021]. At the same time, previous research shows that the clarity and format of energy-use information play a pivotal role in whether individuals adopt meaningful conservation behaviors [@fischerFeedbackHouseholdElectricity2008; @attariPublicPerceptionsEnergy2010; @canfieldPerceptionsElectricityuseCommunications2017]. The interplay of these factors indicates that how energy data is presented can substantially shape not only how people interpret their consumption but also the degree to which they modify their habits to save energy.

@@

Furthermore, research suggests that natural frequencies and absolute numbers are generally easier for individuals to understand compared to percentages or probabilities [@gigerenzerHowImproveBayesian1995 @hoffrageCommunicatingStatisticalInformation2000]. In the context of energy conservation, using absolute units may facilitate more accurate planning and decision-making by aligning with intuitive cognitive processing. However, it's important to note that even intuitive formats can pose challenges. @weberWhyCanOnly2018 found that individuals often struggle with reasoning tasks presented in natural frequencies because they inadvertently revert to more complex probabilistic thinking. This suggests that merely changing the information format may not be sufficient; understanding how consumers process and utilize this information is crucial.

Although prior research shows that energy-use information must be intelligible and action-oriented [@canfieldPerceptionsElectricityuseCommunications2017; @attariPublicPerceptionsEnergy2010], many consumers still have difficulty converting broad guidance (“use less energy”) into targeted appliance-level plans. As highlighted by @abrahamseReviewInterventionStudies2005, information alone, without concrete guidance on how to distribute reductions across appliances or times of use, may fail to encourage meaningful change. Moreover, a meta-analysis by @nematiImpactInformationbasedInterventions2020 revealed that interventions featuring frequent, household-level feedback, particularly when combined with social norm information and monetary incentives, are significantly more effective in driving conservation. Similar findings emerge in contexts beyond electricity. For instance, @tonkeProvidingProceduralKnowledge2024 reports that offering residents a short but actionable list of water-saving strategies—rather than a generic call to conserve—led to a measurable decrease in residential water consumption. These results suggest that equipping consumers with procedure-focused instructions can complement abstract energy-saving goals.

@@

For each family scenario, the participants were shown a table containing the families utility usage from the prior year, alongside the state averages for each appliance category (both prior year usage and state averages are always shown in kWh). For each scenario, participants were asked to create two possible action plans to achieve the target reduction in total household energy usage (see @fig-task). Depending on their reference class condition, the target reduction amount presented either in kilowatt-hours (kWh), as percentages of total household usage, or in U.S. dollars. In all conditions, the target reduction was equivalent to a 15% reduction in total household kWh.

@@

The experimental procedures in Experiment 2 are quite similar to those in Experiment 1. Experiment 2 employed a 2 (task goal: 10% vs. 15% reduction) x 2 (last year's usage: exact vs. rounded) within-subjects design, with a between-subjects manipulation of the reference class (USD vs. Percentage vs. kWh). We recruited 206 participants from Amazon Mechanical Turk, but data from from 10 participants were corrupted due to experimenter error, leaving a final sample of 196 participants.

@@

We once again examined the effect of energy literacy on planning accuracy. A Bayesian linear regression model was fit with log-transformed absolute error as the outcome variable and energy literacy score as the predictor, controlling for random effects of participant and state: log_abs_error ~ els + (1|id) + (1|state). This revealed a significant negative relationship between energy literacy and log absolute error (Estimate = -3.21, 95% CI: -3.89 to -2.52), indicating that participants with higher energy literacy scores tended to have smaller deviations from the target reduction goal, and thus more accurate plans overall (@fig-s2-els).

@@

# General Discussion
